---
title: "Data transformation Using a New York Times API in R"
author: "Pascal Hermann Kouogang Tafo"
format: html
editor: visual
---

## INTRODUCTION

This assignment consists of using the **New York Times Article Search API** to examine how the *volume* and *framing* of soccer coverage in the New York Times has evolved since the modern Major League Soccer (MLS) expansion era began in 2005.In fact d,API provides rich article-level metadata such as headline, publication date, section name, news desk, word count, and multimedia flags since 1851, making it well suited for detecting decade-long window in editorial attention and story framing by querying the keyword **"soccer"**.

------------------------------------------------------------------------

## APPROACH

Below is a structured workflow approach to accomplish our goal and answer our question:

1)  Use a **securely stored environment** variable to **Access the NYT Article Search** and prevent API key exposure

2)  **Filter** and **Collect** the most recent articles containing the keyword "soccer" at request time to ensure a contemporary dataset. The result will contain several fields as nested lists such as "headline", "keywords", "byline\`", "multimedia" etc...

3)  Then i will **parse** each paginated JSON response into a data frame specifically by **extracting only** the sub-fields i need and collapsing arrays to scalar values.

4)  Finally, **clean the data** to calculate which newspaper sections are producing the most content on **"SOCCER"**.

------------------------------------------------------------------------

## Loading the useful libraries

```{r}

library(httr2)       # Modern HTTP client
library(jsonlite)    # JSON parsing
library(tidyverse)   # Data wrangling + ggplot2
library(lubridate)   # Date handling
library(glue)        # String interpolation
library(scales)      # Plot axis formatting
```

------------------------------------------------------------------------

## Let's authenticate our API Key

In this section, i created an account into the New York Times developer portal and i enabled the "Article Search API" and "Most Popular API" to generate an API Key that i securely stored in R through my console to prevent it to be exposed.

```{r}

NYT_API_KEY <- Sys.getenv("NYT_API_KEY")

if (nchar(NYT_API_KEY) == 0) {
  stop(
    "NYT API key not found.\n",
    "  Option A (recommended): add  NYT_API_KEY=your_key  to ~/.Renviron, then restart R.\n",
    "  Option B: assign directly:  NYT_API_KEY <- 'your_key'"
  )
}

BASE_URL <- "https://api.nytimes.com/svc/search/v2/articlesearch.json"

```

------------------------------------------------------------------------

## Let's BUILD THE REQUEST to the Endpoint & Collect DATA

The NYT API provides 10 results per page and maintains a set limit of roughly 12 requests per minute. The script iterates through pages, pausing to remain polite and automatically manage any HTTP 429 rate-limit responses.

```{r}

#' we start by Fetching one page of NYT article search results
#'
#' @param query      Search string
#' @param year       4-digit year (integer)
#' @param page       0-based page index
#' @param api_key    NYT API key string
#' @return           Raw parsed list from the JSON response, or NULL on error

fetch_nyt_page <- function(query, year, page = 0, api_key) {

  begin_date <- glue("{year}0101")
  end_date   <- glue("{year}1231")

  resp <- request(BASE_URL) |>
    req_url_query(
      q          = query,
      begin_date = begin_date,
      end_date   = end_date,
      page       = page,
      sort       = "oldest",
      `api-key`  = api_key
    ) |>
    req_error(is_error = \(r) FALSE) |>   # handle HTTP errors manually
    req_perform()

  if (resp_status(resp) != 200) {
    warning(glue("HTTP {resp_status(resp)} for year={year} page={page}"))
    return(NULL)
  }

  resp |> resp_body_json(simplifyVector = FALSE)
}


#' Now, let's Collect articles across multiple pages for a single year
#'
#' @param query     Search string
#' @param year      4-digit year
#' @param max_pages Max pages to retrieve per year (10 articles each)
#' @param api_key   NYT API key
#' @return          List of article metadata lists


fetch_year <- function(query, year, max_pages = 5, api_key) {

  all_docs <- list()

  for (pg in seq(0, max_pages - 1)) {
    message(glue("  Fetching year={year} page={pg} ..."))
    result <- fetch_nyt_page(query, year, pg, api_key)

    if (is.null(result)) break

    hits  <- result$response$meta$hits %||% 0
    docs  <- result$response$docs

    if (length(docs) == 0) break

    all_docs <- c(all_docs, docs)

    # Stop early if we've retrieved all available articles
    if (length(all_docs) >= hits) break

    Sys.sleep(6.5)   # Stay within 10 req/min rate limit
  }

  all_docs
}

# ── Query parameters ──────────────────────────────────────────────────────────

SEARCH_QUERY <- "soccer MLS"
START_YEAR   <- 2005
END_YEAR     <- as.integer(format(Sys.Date(), "%Y"))
MAX_PAGES    <- 5          # Up to 50 articles per year; increase for deeper pull

message(glue("\nFetching NYT soccer coverage: {START_YEAR}–{END_YEAR}\n"))

raw_articles <- list()

for (yr in START_YEAR:END_YEAR) {
  yr_docs          <- fetch_year(SEARCH_QUERY, yr, MAX_PAGES, NYT_API_KEY)
  raw_articles[[as.character(yr)]] <- yr_docs
  message(glue("  → {length(yr_docs)} articles retrieved for {yr}"))
  Sys.sleep(2)
}


```

## LET'S PARSE THE JSON RESPONSE

Since each article doc is a nested list, we will extract the fields we need.

```{r PARSE-JSON-RESPONSE}

# Safely extract a nested field (returns NA if missing)

safe_pluck <- function(x, ...) {
  tryCatch(purrr::pluck(x, ...) %||% NA_character_, error = \(e) NA_character_)
}

parse_article <- function(doc) {
  tibble(
    article_id    = safe_pluck(doc, "_id"),
    pub_date      = safe_pluck(doc, "pub_date"),
    headline      = safe_pluck(doc, "headline", "main"),
    abstract      = safe_pluck(doc, "abstract"),
    lead_paragraph= safe_pluck(doc, "lead_paragraph"),
    section       = safe_pluck(doc, "section_name"),
    desk          = safe_pluck(doc, "news_desk"),
    doc_type      = safe_pluck(doc, "document_type"),
    word_count    = as.integer(safe_pluck(doc, "word_count") %||% NA),
    web_url       = safe_pluck(doc, "web_url"),
    # Combine all keywords into a semicolon-separated string
    keywords      = {
      kws <- doc$keywords
      if (length(kws) > 0) {
        paste(sapply(kws, \(k) k$value %||% ""), collapse = "; ")
      } else NA_character_
    }
  )
}

```

------------------------------------------------------------------------

## LET'S TRANSFORM THE JSON RESPONSE INTO A CLEAN DATA FRAME

```{r}

# Step 1: Flatten year buckets into a single list of docs

articles_df <- raw_articles |>
  purrr::list_flatten() |>
  purrr::map(parse_article) |>
  purrr::list_rbind() |>

# Step 2: Let's transform raw date strings into structured time components.
  mutate(
    pub_date   = ymd_hms(pub_date, quiet = TRUE),
    year       = year(pub_date),
    month      = month(pub_date),
    quarter    = quarter(pub_date),
    year_month = floor_date(pub_date, "month")
  ) |>

  # Step 3: Framing (These classify each article's emphasis based on headline + abstract text.)
  
  mutate(
    text_combined = str_to_lower(paste(headline, abstract, lead_paragraph, sep = " ")),

    frame_growth      = str_detect(text_combined,
                          "expansion|grow|record|attendance|stadium|franchise|invest"),
    frame_mls         = str_detect(text_combined, "\\bmls\\b|major league soccer"),
    frame_usmnt       = str_detect(text_combined,
                          "u\\.s\\. national|usmnt|uswnt|world cup|pulisic|reyna"),
    frame_global      = str_detect(text_combined,
                          "premier league|champions league|barcelona|real madrid|messi|ronaldo"),
    frame_transfer    = str_detect(text_combined,
                          "transfer|signing|contract|deal|salary|designated player"),
    frame_youth       = str_detect(text_combined,
                          "youth|academy|college|high school|young"),
    frame_culture     = str_detect(text_combined,
                          "culture|fan|supporter|diverse|immigrant|latino|community"),

# Step 4: perform conditional labeling (first match wins; "Other" if none)

    framing = case_when(
      frame_mls      ~ "MLS / League",
      frame_usmnt    ~ "US National Team",
      frame_global   ~ "Global / Club",
      frame_transfer ~ "Transfers & Business",
      frame_growth   ~ "Growth & Infrastructure",
      frame_youth    ~ "Youth & Development",
      frame_culture  ~ "Culture & Fandom",
      TRUE           ~ "Other / General"
    )
  ) |>

# Step 5: Let's Deduplicate the process and clean the data
  
  distinct(article_id, .keep_all = TRUE) |>
  filter(!is.na(pub_date), year >= START_YEAR) |>
  arrange(pub_date)

message(glue("\nFinal dataset: {nrow(articles_df)} unique articles ({START_YEAR}–{END_YEAR})"))

head(articles_df,10)

```

## As a result, our dataframe contains 142 unique articles of 142 rows and 24 columns including article_id, pulication date, headline, doc_type, etc...

## ANALYSIS & VISUALISATIONS

### 1. Annual article volume

Here, we will analyze and visualize the NYT Soccer Coverage Volume since MLS Expansion Era.

```{r}

annual_volume <- articles_df |>
  count(year, name = "n_articles")

annual_volume

ggplot(annual_volume, aes(x = year, y = n_articles)) +
  geom_col(fill = "#003087", alpha = 0.85) +
  geom_smooth(method = "loess", se = TRUE, colour = "#E31837", linewidth = 1) +
  scale_x_continuous(breaks = seq(START_YEAR, END_YEAR, 2)) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "NYT Soccer Coverage Volume: MLS Expansion Era",
    subtitle = glue("Articles matching '{SEARCH_QUERY}', {START_YEAR}–{END_YEAR}"),
    x = NULL, y = "Number of Articles",
    caption  = "Source: NYT Article Search API"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))


```

### 2) Framing breakdown by year

```{r}
framing_year <- articles_df |>
  count(year, framing) |>
  group_by(year) |>
  mutate(pct = n / sum(n))
framing_year

p2 <- ggplot(framing_year, aes(x = year, y = pct, fill = framing)) +
  geom_area(alpha = 0.85, colour = "white", linewidth = 0.3) +
  scale_x_continuous(breaks = seq(START_YEAR, END_YEAR, 2)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "How NYT Soccer Coverage is Framed Over Time",
    subtitle = "Share of articles by dominant frame, per year",
    x = NULL, y = "Share of Articles", fill = "Frame",
    caption  = "Source: NYT Article Search API"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title   = element_text(face = "bold"),
    legend.position = "right"
  )

print(p2)
```
